\vspace{24pt} Comparison of Different Data Layouts for Columnar In-Memory Databases on the Basis of an Example Application

### 2018-08-31 paper/paper.tex
sime typos

\maketitle
\begin{abstract}
With are more and more devices that are equipped with GPS and generate trajectory data analyzing and using this data catches attention increasingly. That is why besides using some specialized trajectory processing systems for that kind of data the interest in using common databases, especially columnar in-memmory systems, for this task rises. We compare three different data layouts for such databases based on a real-world usecase. Therefore we developed an example application with taxi data and benchmarked our those data layouts against typical queries of our usecase.
\end{abstract}

\keywords{Geospatial data, Trajectory analysis, Taxi data, In-Memory}

@@

Storing trajectory data in traditional relational databases is not always the best idea. Requests might take a long time when working on big data sets. It is also difficult to achieve good compression rates. There exist some systems which are purely designed to work with such trajectories. Those systems are performing much better than the traditional databases.

In our project we would like to investiage how a columnar in-memory database can handle trajectory data. Thereby, we can make use of the in-memory architecture and of the columnar storage benefits. We developed an example application which covers a lot of real world requests to a large trajectory data set. This application builds a good basis to benchmark different data layouts and optimizations.

Traditional taxi companies like to increase their profit. For that, the individual taxi drivers would have to find many passengers and minimize their waiting times. A way to be a  good performing taxi is to stay in a profitable area of the city. For example, where taxi rides are highly needed at a specific time. With our help, the taxi companies can develop new strategies. A simple strategy would be to only accept and assign orders whose destination is in a good area for the expected arrival time. We created an interactive map, which shows the best performing taxis for one day in Shenzhen. This is based on a data set that is explained in section \ref{sec:ds}. The application has some additional information about the taxi density and numbers about pick-ups and drop-offs. Taxi companies could use the application to quickly analyze the collected GPS data.\\

@@

The data we used is a freely available collection of GPS information of taxis.\footnote{\href{https://www-users.cs.umn.edu/~tianhe/BIGDATA/UrbanCPS/TaxiData/TaxiData}{https://www-users.cs.umn.edu/$\sim$tianhe/BIGDATA/\\UrbanCPS/TaxiData/TaxiData}} The dataset set has been collected over one day starting at 12~AM and ending at 23:59~PM in the city of Shenzhen, China and the surrounding area. The data does not only contain the GPS location at a certain time, but also the current speed and the taxi's occupancy. The status of a driving taxi has been recorded in an average interval of 26.85 seconds or every 15 seconds in median. In total there are 14,728 trajectories traced resulting in almost 47 million records. The data comes in CSV-format which can be loaded into the SAP HANA database immediately. One sample point contains six attributes: trajectory id (TID), longitude, latitude, timestamp, speed and a flag indicating the presence of a passenger.

\subsection{Data Cleansing}
Before we started analyzing the trajectories we had a closer look at the dataset itself. We noticed rather soon, that there are some inconsistencies within data. Hence, we perform data cleansing. In the following paragraphs, we describe our criteria for data cleansing in detail.

\subsubsection{Duplicates}
The data was already imported into the database by our supervisor, when we started our project. Since we did not want to change the original data, we created an own table. Once we wanted to create a primary key, the database has thrown an error stating that the data has duplicate values. In fact, there are 381 data points occuring not only twice, but also multiple times. We deleted those entries, because they do not add any value to the data and prevent the creation of a primary key.

\subsubsection{Bounding box}
Generally, GPS coordinates are not completely accurate. However, there are points in the data set, that do not make sense, are not even possible or do not fit into the geospatial system of longitudes and latidues. The graphic \ref{fig:bbox} below shows the trajectory with id \textit{22,360} which passes through the ocean. In order to eliminate such outliers, we use a bounding box (rectangle highlighted in red color). To avoid deleting points that are outside this area, because there a taxi really drove there, we have chosen the rectangle intentionally large. This data cleansing step effects almost 8,000 GPS points.

@@

The backend essentially has a REST API which is written in Python and uses the Flask framework. It serves the static HTML and Javascript files, as well as redirecting requests to the database. The backend also has a few python scripts that are used to create tables and for the data cleansing.

@@

Since the data neither changes nor grows anymore, we decided to store the information in sorted order. As sorting key we utilized the trajectory identifier and the timestamp. In order to speed up window queries we materialized a unique identifier, which is the row ID. This ID simplifies the creation of a primary key. Otherwise, a primary key would have to include multiple columns, most likely all of them. The Hana database automatically creates an index on this column. Furthermore, we realized that timestamp operations perform relatively bad. In order to avoid timestamps we calculate the amount of seconds lasted and store those as an integer in the database.

@@

\subsubsection{Get whole trajectory}
In various situations we need to have a whole in hand. Therefore, we wrote SQL queries for each storage layout. The query execution times are shown in graphic \ref{fig:bench_whole_trajectory}. In order to create the statistics we ran the queries 100 times. The time is measured within our python server and encoompasses the whole recreation. Network latency between the server and the database has not been measured. Hence, it is not part of the benchmark.

The slowest query execution time feature the sample point format. However, the data retrieved from the database does not need to be recreated in the python backend. It can be used as it is.

Recreating a trajectory in the frame format is faster in terms of retrieving the data from the database. However, in order to calculate the correct GPS coordinates there is a computation overhead. We have done this in the python backend. An attempt of recreating the data via an SQL procedure using dynamic SQL failed due to extraordinary execution times. This takes 89.3 seconds, which is a performance decrease by factor 4000. An explanation for the high number could be, that dynamic SQL cannot be optimized well by the DBMS and runs in a single threaded style. The frame layout having a frame size of 15 seconds almost takes double the time as the frame size of 30 seconds. The reason for that, is the double amount of data.

@@

To calculate the distance we demand to do it completely in the database. When querying a full trajectory we have to use the sample point format, since any other layout depends on the python backend.

The simplest way to get the distance of a ride is the usage of spatial points (\textit{ST\_POINTS}). This is a data type implemented in SAP Hana, that can be created out of a longitude and a latitude value. Furthermore, there is a function to compute the distance between those points. A single ride starts, when a passenger enters a taxi (occupancy changes from zero to one) and ends, when the passenger leaves (occupancy changes from one to zero). We sum up the lengths between all points sum uo all values for a whole trajectory.

>>>>

\maketitle
\begin{abstract}
With are more and more devices that are equipped with GPS and generate trajectory data analyzing and using this data catches attention increasingly. That is why besides using some specialized trajectory processing systems for that kind of data the interest in using common databases, especially columnar in-memmory systems, for this task rises. We compare three different data layouts for such databases based on a real-world use case. Therefore we developed an example application with taxi data and benchmarked our those data layouts against typical queries of our use case.
\end{abstract}

% looks awkward without line break
\keywords{Geospatial data, Trajectory analysis, Taxi data,\\In-Memory}

@@

Storing trajectory data in traditional relational da\-ta\-ba\-ses is not always the best idea. Requests might take a long time when working on big datasets. It is also difficult to achieve good compression rates. There exist some systems which are purely designed to work with such trajectories. Those systems are performing much better than the traditional databases.

In our project investigated how a columnar in-memory database can handle trajectory data. Thereby, we can make use of the in-memory architecture and the columnar storage benefits. We developed an example application which covers a lot of real world requests to a large trajectory dataset. This application builds a good basis to benchmark different data layouts and optimizations.

Traditional taxi companies like to increase their profit. For that, the individual taxi drivers would have to find many passengers and minimize their waiting times. A way to be a good performing taxi is to stay in a profitable area of the city. For example, where taxi rides are highly needed at a specific time. With our help, the taxi companies can develop new strategies. A simple strategy would be to only accept and assign orders whose destination is in a good area for the expected arrival time. We created an interactive map, which shows the best performing taxis for one day in Shenzhen. This is based on a dataset that is explained in section \ref{sec:ds}. The application has some additional information about the taxi density and numbers about pick-ups and drop-offs. Taxi companies could use the application to quickly analyze the collected GPS data.

@@

The data we used is a freely available collection of GPS information of taxis.\footnote{\href{https://www-users.cs.umn.edu/~tianhe/BIGDATA/UrbanCPS/TaxiData/TaxiData}{https://www-users.cs.umn.edu/$\sim$tianhe/BIGDATA/\\UrbanCPS/TaxiData/TaxiData}} The dataset has been collected over one day starting at 12~AM and ending at 23:59~PM in the city of Shenzhen, China and the surrounding area. The data does not only contain the GPS location at a certain time, but also the current speed and the taxi's occupancy. The status of a driving taxi has been recorded in an average interval of 26.85 seconds or every 15 seconds in median. In total there are 14,728 trajectories traced resulting in almost 47 million records. The data comes in CSV-format which can be loaded into the SAP HANA database immediately. One sample point contains six attributes: trajectory id (TID), longitude, latitude, timestamp, speed and a flag indicating the presence of a passenger.

\subsection{Data Cleansing}
Before we started analyzing the trajectories we had a closer look at the dataset itself. We noticed rather soon, that there are some inconsistencies within the data. Hence, we perform data cleansing. In the following paragraphs, we describe our criteria for data cleansing in detail.

\subsubsection{Duplicates}
The data was already imported into the database by our supervisor, when we started our project. Since we did not want to change the original data, we created an own table. Once we wanted to create a primary key on it, the database has thrown an error stating that the data has duplicate values. In fact, there are 381 data points occuring not only twice, but also multiple times. We deleted those entries, because they do not add any value to the data and prevent the creation of a primary key.

\subsubsection{Bounding box}
Generally, GPS coordinates are not completely accurate. However, there are points in the dataset, that do not make sense, are not even possible or do not fit into the geospatial system of longitudes and latitudes. The graphic \ref{fig:bbox} below shows the trajectory with id \textit{22,360} which passes through the ocean. In order to eliminate such outliers, we use a bounding box (rectangle highlighted in red color). To avoid deleting points that are outside this area, because a taxi really drove there, we have chosen the rectangle intentionally large. This data cleansing step effects almost 8,000 GPS points.

@@

The backend essentially has a REST API which is written in Python and uses the Flask framework. It serves the static HTML and Javascript files, as well as redirecting requests to the database. The backend also has a few Python scripts that are used to create tables and for the data cleansing.

@@

Since the data neither changes nor grows anymore, we decided to store the information in sorted order. As sorting key we utilized the trajectory identifier and the timestamp. In order to speed up window queries we materialized a unique identifier, which is the row ID. This ID simplifies the creation of a primary key. Otherwise, a primary key would have to include multiple columns, most likely all of them. The HANA database automatically creates an index on this column. Furthermore, we realized that timestamp operations perform relatively bad. In order to avoid timestamps we calculate the amount of seconds lasted and store those as an integer in the database.

@@

\subsubsection{Get whole trajectory}
In various situations we need to have a whole in hand. Therefore, we wrote SQL queries for each storage layout. The query execution times are shown in graphic \ref{fig:bench_whole_trajectory}. In order to create the statistics we ran the queries 100 times. The time is measured within our Python server and encoompasses the whole recreation. Network latency between the server and the database has not been measured. Hence, it is not part of the benchmark.

The slowest query execution time feature the sample point format. However, the data retrieved from the database does not need to be recreated in the Python backend. It can be used as it is.

Recreating a trajectory in the frame format is faster in terms of retrieving the data from the database. However, in order to calculate the correct GPS coordinates there is a computation overhead. We have done this in the Python backend. An attempt of recreating the data via an SQL procedure using dynamic SQL failed due to extraordinary execution times. This takes 89.3 seconds, which is a performance decrease by factor 4000. An explanation for the high number could be, that dynamic SQL cannot be optimized well by the DBMS and runs in a single threaded style. The frame layout having a frame size of 15 seconds almost takes double the time as the frame size of 30 seconds. The reason for that, is the double amount of data.

@@

To calculate the distance we demand to do it completely in the database. When querying a full trajectory we have to use the sample point format, since any other layout depends on the Python backend.

The simplest way to get the distance of a ride is the usage of spatial points (\textit{ST\_POINTS}). This is a data type implemented in SAP HANA, that can be created out of a longitude and a latitude value. Furthermore, there is a function to compute the distance between those points. A single ride starts, when a passenger enters a taxi (occupancy changes from zero to one) and ends, when the passenger leaves (occupancy changes from one to zero). We sum up the lengths between all points sum uo all values for a whole trajectory.
